---
title: "01_data_cleaning"
format: pdf
editor: visual
---

## Overview of File

This file prepares the dataset for analysis. The file converts the raw dataset into a clean and usable format for exploratory data analysis and Bayesian modeling.

The cleaning process includes:

-   loading the dataset

-   inspecting the structure of the raw data

-   renaming variables to simpler names

-   converting variables to consistent binary formats

## Data Cleaning

```{r}
#| include: false
# Load Required Libraries
library(tidyverse)
library(janitor)
```

```{r}
# read in the raw dataset
dataset <- read.csv("../data/data.csv")
```

```{r}
# inspect structure of the raw dataset
glimpse(dataset)
```

```{r}
# renaming variables to simpler names
data <- dataset |>
  select(
    gallstone = Gallstone.Status,
    gender = Gender,
    age = Age,
    bfr = Total.Body.Fat.Ratio..TBFR.....,
    bmi = Body.Mass.Index..BMI.,
    glucose = Glucose,
    comorbid = Comorbidity,
    height = Height,
    weight = Weight,
    tbw = Total.Body.Water..TBW.,
    ecw = Extracellular.Water..ECW.,
    icw = Intracellular.Water..ICW.,
    ecftbw = Extracellular.Fluid.Total.Body.Water..ECF.TBW.,
    tbfr = Total.Body.Fat.Ratio..TBFR.....,
    lean_mass = Lean.Mass..LM.....,
    bpc = Body.Protein.Content..Protein.....,
    vfr = Visceral.Fat.Rating..VFR.,
    bone_mass = Bone.Mass..BM.,
    musc_mass = Muscle.Mass..MM.,
    obesity = Obesity....,
    tfc = Total.Fat.Content..TFC.,
    vfa = Visceral.Fat.Area..VFA.,
    vma = Visceral.Muscle.Area..VMA...Kg.,
    hfa = Hepatic.Fat.Accumulation..HFA.,
    tc = Total.Cholesterol..TC.,
    ldl = Low.Density.Lipoprotein..LDL.,
    hdl = High.Density.Lipoprotein..HDL.,
    trigly = Triglyceride,
    alt = Alanin.Aminotransferaz..ALT.,
    alp = Alkaline.Phosphatase..ALP.,
    creat = Creatinine,
    gfr = Glomerular.Filtration.Rate..GFR.,
    crp = C.Reactive.Protein..CRP.,
    hgb = Hemoglobin..HGB.,
    vita_d = Vitamin.D
  ) 
```

```{r}
# reverse coding of gallstone variable
data <- data |>
  mutate(gallstone = ifelse(gallstone == 0, 1, 0))
```

The gallstone variable indicates if a patient has gallstones. We convert the values so that the variable is coded as a binary outcome:

-   1 = patient has gallstones

-   0 = patient does not have gallstones

In the original dataset, the value 0 indicated that a patient had gallstones. We reversed this coding so that 1 represents having gallstones and 0 represents not having gallstones. This is important because logistic regression models have the probability of the outcome coded as 1, so we want the value 1 to represent the event of interest, which in this case is having gallstones.

```{r}
# convert comorbid variable to binary
data <- data |>
  mutate(comorbid = ifelse(comorbid == 0, 0, 1))
```

In the original dataset, the comorbid variable contained multiple categories representing different levels of comorbidity. For this analysis, we converted it into a binary variable where 0 indicates no comorbidities and 1 indicates the presence of at least one comorbidity. This simplifies the variable and allows us to examine whether having any comorbidity is associated with gallstones without modeling several separate categories.

```{r}
# save cleaned dataset so it can be used in later analysis files
write.csv(data, "../data/cleaned_data.csv", row.names = FALSE)

```
